Use and acquisition of causal relations for decision making under uncertainty using imperfect information repeated games

### 2018-06-12 Propuesta/Proposal_English.tex
teo de juegos: normal form games

@@

For what remains of this section, we follow \cite{bernardo2000bayesian} in his exposition of Savage's theory since it is the most suitable for our purposes because we will be considering decision makers that do not know all of their environment.

	\subsubsection{Elements of a Decision Problem}
	A Decision Problem under Uncertainty is composed by:
	\begin{itemize}
	\item A set $\Omega$ of \textit{states}.
	\item A set $\mathcal{A}=\{a_i : i \in I \}$ of actions.
	\item For each action $a_i$, a partition $E_i = \{ E_j : j \in J \}$ of $\Omega$. Define $\mathcal{E} = \cup_{i \in I } E_i$.
	\item For each action $a_i$, a set of consequences $C_i = \{ c_j : j \in J \}$. Define $\mathcal{C}=\cup_{i \in I} C_i$
	\item A preference relation $\succeq$ defined over $\mathcal{A}$ which represent the decision-maker whishes. It is assumed that the decision maker will choose among his options according to his preferences, which are represented by $\succeq$.
	\end{itemize}
	Since we are considering that a decision maker chooses an action $a \in \mathcal{A}$, and then an uncertain event $E$ will occur which will then produce the consequence $c$, we will identify actions in $\mathcal{A}$ as $a_i = \{ c_j | E_j : j \in J \}$. As a technical assumption required for all of the math to work, we require that elements of $\mathcal{C}$ belong to $\mathcal{A}$. This is achieved by adding into $A$ elements of the form $\{ c | \Omega \}$ for $c \in \mathcal{C}$. This being said, we can extend the preference relation $\succeq$ to elements of $\mathcal{C}$ and therefore indistinctively write $c_1 \succeq c_2$ for $c_1, c_2 \in \mathcal{C}$ or $a_1 \succeq a_2$ for $a_1,a_2 \in \mathcal{A}$. We require that $\mathcal{E}$ is an \textit{algebra}.	
	\begin{defi}
	From the prerefence relation $\succeq$ we can derive some other relations between actions:
	\begin{itemize}
	\item $a_1 \sim a_2 \Leftrightarrow a_1 \succeq a_2 \textrm{ and } a_2 \succeq a_1$
	\item $a_1 \succ a_2 \Leftrightarrow a_1 \succeq a_2 \textrm{ and it does not hold that } a_2 \succeq a_1$.
	\end{itemize}
	\end{defi}
	\begin{defi}
	A relation between events can be derived from $\succeq$ in the following way:
	\[E \succeq F \Leftrightarrow \textrm{ for all } c_1 \succeq c_2 \textrm{ it holds that } \{ c_2 | E, c_1 | E^c \} \succeq \{ c_2 | F, c_1 | F^c \}. \]
	In this case, we say that $F$ is not \textit{more likely} than $E$. It can easily be shown that $\Omega \succ \emptyset$, and a relation $\sim$ for events is defined in an analog way than for consequences.
	\end{defi}
	\begin{defi}
	Given some $G \succ \emptyset$ we define a conditional preference relation as follows:
	\begin{itemize}
	\item i) $a_1 \succeq_G a_2 \Leftrightarrow \textrm{ for all a } \{ a_1 | G, a | G^c \} \succeq \{ a_2 | G, a | G^c \}$.
	\item ii) $E \succeq_G F \Leftrightarrow \textrm{ for } c_1 \succeq_G c_2 \{ c_2 | E , c_1 | E^c \} \succeq_G \{c_2 | F , c_1  | F^c \}$ 
	\end{itemize}
	\end{defi}
	\begin{defi}
	Two events $E,F \in \mathcal{E}$ are said to be independent if and only if:
	\begin{itemize}
	\item 	
	\end{itemize}
	\end{defi}
	\subsubsection{Axioms of Coherence}
	We require the following axioms for the preference relation $\succeq$ which we will call Axioms of Coherence or Rationality, and define a Rational decision maker as a decision maker who will choose according to a preference relation and whose preferences are rational in the sense of the axioms (\cite{bernardo2000bayesian}).\\
	\\
	\indent \textbf{Axiom 1}.
	\begin{itemize}
	\item i) There exists consequences $c_1$, $c_2$ such that $c_1 \succ c_2$.
	\item ii) For all consequences $c_1, c_2$ and events $E,F \in \mathcal{E}$ either $\{  c_2 | E , c_1 | E^c\} \succeq \{ c_2 | F, c_1 | F^c \}$ or $\{ c_2 | F, c_1 | F^c \} \succeq \{  c_2 | E , c_1 | E^c\}$
	\end{itemize}

@@

	For interpretations of the axioms in diverse contexts we refer the reader to \cite{bernardo2000bayesian}, \cite{binmore2008rational}, \cite{gilboa2009decision}, \cite{wakker2010prospect}.
	\subsubsection{Subjective Probability and Utility}
	\begin{defi}
	Given a preference relation $\succeq$, we define the (subjective) \textbf{probability} of an event $E \in \mathcal{E}$ as the real number $\mu(S)$ associated with the standard event $S$ such that $E \sim S$.
	\end{defi}
	The subjective probability thus defined satisfies all of the Kolmogorov's axioms of probability (\cite{bernardo2000bayesian}). So, we can use all of the mathematical machinery for classic probability measures, including the definition of conditional probability, Bayes' theorem, etc.
	\begin{defi}
	Consequences $c^\ast$ and $c_\ast$ are called respectively best and worst (extreme consequences) if for any other consequence $c \in \mathcal{C}$ then $c^\ast \succeq c \succeq c_\ast$. Decision problems in which we add extreme consequences are called bounded decision problems.
	\end{defi}
	\begin{defi}
	Given a preference relation $\succeq$ we define the canonical \textbf{utility} $u(c)=u(c | c_\ast, c^\ast)$ of a consequence $c \in \mathcal{C}$ relative to the extreme consequences $c_\ast$, $c^\ast$, as the real number $\mu(S)$ associated with any standard event $S$ such that $c \sim \{c^\ast | S, c_\ast | S^c\}$.
	\end{defi}
	\subsubsection{Decision Criteria for Decision Problems}
	\begin{prop}
	For any bounded decision problem with extreme consequences $c_\ast$ and $c^\ast$ it holds:
	\begin{itemize}
	\item i) For all $c \in \mathcal{C}$, $u(c | c_\ast, c^\ast)$ exists and is unique.
	\item ii) The value of $u(c | c_\ast, c^\ast)$ is unafected by the occurrence of any event $G \succ \emptyset$.
	\item iii) $0 = u(c_\ast | c_\ast, c^\ast) \leq u(c | c_\ast, c^\ast) \leq u(c^\ast | c_\ast, c^\ast) = 1$.
	\end{itemize}
	\end{prop}
	\begin{defi}
	For a bounded decision problem and any event $G \succ \emptyset$ and $a = \{ c_j | E_j : j \in J \}$ we define the conditional expected utility of $a$ as
	\[ \bar{u}(a | c_\ast, c^\ast, G) = \sum_{j \in J} u(c_j | c_\ast, c^\ast)P(E_j | G). \]
	If $G = \Omega$ we simply denote $\bar{u}(a | c_\ast, c^\ast, G)$ as $\bar{u}(a | c_\ast, c^\ast)$
	\end{defi}
	\begin{teo}
	For any bounded decision problem with extreme consequences $c^\ast \succ c_\ast$ and any $G \succ \emptyset$
	\[ a_1 \succeq_G a_2 \Leftrightarrow \bar{u}(a_1) \geq \bar{u}(a_2). \]
	\end{teo}
	\begin{proof}
	See \cite{bernardo2000bayesian} Chapter 2 Proposition 2.22
	\end{proof}
	This result means that for any decision maker whose preferences satisfies axioms 1 through 5 the only election criteria that is compatible with the axioms is the maximization of expected utility. Of course, this result establishes a complete ordering of the options, but it does not guarantees the existence of an option for which the expected utility is maximum (\cite{bernardo2000bayesian}).
	\subsection{Game Theory}

\section{Problem Statement}
	\subsection{Problem Statement}
	\subsection{Motivation}
	Decision making under uncertain conditions is a fundamental part of intelligent reasoning (\cite{lake2017building}). Intelligent agents often face situations where an action must be chosen in the presence of uncertain conditions; this means that an outcome will be observed according to some probability distribution given the action chosen by the agent.

@@

	It is known that human beings conceive their actions on the world as \textit{intervening} in the world. Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem as if choosing an intervention was choosing an arm of a \textit{slot machine}, in which a gambler chooses an arm and gets some reward. From the rewardsm they estimate probabilities and output an optimal action in the sense of obtaining minimal regret.

	 Their work considers that the causal model is fully known. They mention that the case where the causal model is unknown is left as an open question, and it is precisely what we are proposing to answer. 
	\subsection{Justification}
	Besides obtaining a criteria of choosing actions, our agent will also learn a causal model of his environment. Learning causal models is importante since it allows further questions to be asked. 
	\subsection{Research Question}
	\subsection{Hypothesis}
	Let $\mathcal{G}$ a causal graphical model whose probability distribution $P_\mathcal{G}$ satisfies (the axioms). Let $(D,E,c)$ a decision problem whose uncertain events are related to the variables in $\mathcal{G}$. Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$,  by repeatedly making decisions he can learn causal information and use it to find the optimal actions (in the sense of expected utility theory) in less rounds than if he doesn't consider causal information.
	\subsection{Objectives}
		\subsubsection{General Objectives}
		\subsubsection{Specific Objectives}
		\subsubsection{Limitations}

@@

\section{Methodology}
\subsection{Modelling}
\subsection{Working plan}

\newpage
\bibliographystyle{apalike}
\bibliography{/Users/MauricioGS1/INAOE/Propuesta/Bibliografia.bib}
\end{document}

>>>>

	A causal graphical model (CGM) consists of a set of random variables $\mathcal{X}=\{ X_1,...,X_n \}$, an acyclic directed graph (DAG) whose nodes are in correspondence with the variables in $\mathcal{X}$ and whose edges represent relations of cause-effect. Also, the model, which up to this point is nothing more than a Bayesian Network with causal semantics, is enriched with an operator named $do()$ which is defined over graphs and whose action is described as follows:

@@

For what remains of this section, we follow \cite{bernardo2000bayesian} in his exposition of Savage's theory since it is the most suitable for our purposes because we will be considering decision makers that do not know all of their environment. Savage's Decision Making theory, together with von Neumann's theory is what it is commonly known as \textit{classical} decision making.

	\subsubsection{Elements of a Decision Problem}
	A Decision Problem under Uncertainty is composed by:
	\begin{itemize}
	\item A set $\Omega$ of \textit{states}.
	\item A set $\mathcal{A}=\{a_i : i \in I \}$ of actions.
	\item For each action $a_i$, a partition $E_i = \{ E_j : j \in J \}$ of $\Omega$. Define $\mathcal{E} = \cup_{i \in I } E_i$.
	\item For each action $a_i$, a set of consequences $C_i = \{ c_j : j \in J \}$. Define $\mathcal{C}=\cup_{i \in I} C_i$
	\item A preference relation $\succeq$ defined over $\mathcal{A}$ which represent the decision-maker whishes. It is assumed that the decision maker will choose among his options according to his preferences, which are represented by $\succeq$.
	\end{itemize}
	Since we are considering that a decision maker chooses an action $a \in \mathcal{A}$, and then an uncertain event $E$ will occur which will then produce the consequence $c$, we will identify actions in $\mathcal{A}$ as $a_i = \{ c_j | E_j : j \in J \}$. As a technical assumption required for all of the math to work, we require that elements of $\mathcal{C}$ belong to $\mathcal{A}$. This is achieved by adding into $A$ elements of the form $\{ c | \Omega \}$ for $c \in \mathcal{C}$. This being said, we can extend the preference relation $\succeq$ to elements of $\mathcal{C}$ and therefore indistinctively write $c_1 \succeq c_2$ for $c_1, c_2 \in \mathcal{C}$ or $a_1 \succeq a_2$ for $a_1,a_2 \in \mathcal{A}$. We require that $\mathcal{E}$ is an \textit{algebra}.	
	\begin{defi}
	From the prerefence relation $\succeq$ we can derive some other relations between actions:
	\begin{itemize}
	\item $a_1 \sim a_2 \Leftrightarrow a_1 \succeq a_2 \textrm{ and } a_2 \succeq a_1$
	\item $a_1 \succ a_2 \Leftrightarrow a_1 \succeq a_2 \textrm{ and it does not hold that } a_2 \succeq a_1$.
	\end{itemize}
	\end{defi}
	\begin{defi}
	A relation between events can be derived from $\succeq$ in the following way:
	\[E \succeq F \Leftrightarrow \textrm{ for all } c_1 \succeq c_2 \textrm{ it holds that } \{ c_2 | E, c_1 | E^c \} \succeq \{ c_2 | F, c_1 | F^c \}. \]
	In this case, we say that $F$ is \textbf{not more likely} than $E$. It can easily be shown that $\Omega \succ \emptyset$, and a relation $\sim$ for events is defined in an analog way than for consequences.
	\end{defi}
	\begin{defi}
	Given some $G \succ \emptyset$ we define a conditional preference relation $\succeq_G$ as follows:
	\begin{itemize}
	\item i) $a_1 \succeq_G a_2 \Leftrightarrow \textrm{ for all a } \{ a_1 | G, a | G^c \} \succeq \{ a_2 | G, a | G^c \}$.
	\item ii) $E \succeq_G F \Leftrightarrow \textrm{ for } c_1 \succeq_G c_2 \{ c_2 | E , c_1 | E^c \} \succeq_G \{c_2 | F , c_1  | F^c \}$ 
	\end{itemize}
	\end{defi}
	\begin{defi}
	Two events $E,F \in \mathcal{E}$ are said to be independent if and only if:
	\begin{itemize}
	\item 	
	\end{itemize}
	\end{defi}
	\subsubsection{Axioms of Coherence}
	We require the following axioms for the preference relation $\succeq$ which we will call Axioms of Coherence or Rationality, and define a Rational decision maker as a decision maker who will choose according to a preference relation and whose preferences satisfy the axioms (\cite{bernardo2000bayesian}, \cite{gilboa2009decision}).\\
	\\
	\indent \textbf{Axiom 1}.
	\begin{itemize}
	\item i) There exists consequences $c_1$, $c_2$ such that $c_1 \succ c_2$.
	\item ii) For all consequences $c_1, c_2$ and events $E,F \in \mathcal{E}$ either $\{  c_2 | E , c_1 | E^c\} \succeq \{ c_2 | F, c_1 | F^c \}$ or $\{ c_2 | F, c_1 | F^c \} \succeq \{  c_2 | E , c_1 | E^c\}$
	\end{itemize}

@@

	For interpretations of the axioms in diverse contexts we refer the reader to \cite{bernardo2000bayesian}, \cite{binmore2008rational}, \cite{gilboa2009decision}, \cite{wakker2010prospect}.
	\subsubsection{Subjective Probability and Utility}
	\begin{defi}
	Given a preference relation $\succeq$, we define the (subjective) \textbf{probability} of an event $E \in \mathcal{E}$ as the real number $\mu(S)$ associated with the standard event $S$ such that $E \sim S$.
	\end{defi}
	The subjective probability thus defined satisfies all of the Kolmogorov's axioms of probability (\cite{bernardo2000bayesian}). So, we can use all of the mathematical machinery for classic probability measures, including the definition of conditional probability, Bayes' theorem, etc.
	\begin{defi}
	Consequences $c^\ast$ and $c_\ast$ are called respectively best and worst (extreme consequences) if for any other consequence $c \in \mathcal{C}$ then $c^\ast \succeq c \succeq c_\ast$. Decision problems in which we add extreme consequences are called bounded decision problems.
	\end{defi}
	\begin{defi}
	Given a preference relation $\succeq$ we define the canonical \textbf{utility} $u(c)=u(c | c_\ast, c^\ast)$ of a consequence $c \in \mathcal{C}$ relative to the extreme consequences $c_\ast$, $c^\ast$, as the real number $\mu(S)$ associated with any standard event $S$ such that $c \sim \{c^\ast | S, c_\ast | S^c\}$.
	\end{defi}
	\subsubsection{Decision Criteria for Decision Problems}
	\begin{prop}
	For any bounded decision problem with extreme consequences $c_\ast$ and $c^\ast$ it holds:
	\begin{itemize}
	\item i) For all $c \in \mathcal{C}$, $u(c | c_\ast, c^\ast)$ exists and is unique.
	\item ii) The value of $u(c | c_\ast, c^\ast)$ is unafected by the occurrence of any event $G \succ \emptyset$.
	\item iii) $0 = u(c_\ast | c_\ast, c^\ast) \leq u(c | c_\ast, c^\ast) \leq u(c^\ast | c_\ast, c^\ast) = 1$.
	\end{itemize}
	\end{prop}
	\begin{defi}
	For a bounded decision problem and any event $G \succ \emptyset$ and $a = \{ c_j | E_j : j \in J \}$ we define the conditional expected utility of $a$ as
	\[ \bar{u}(a | c_\ast, c^\ast, G) = \sum_{j \in J} u(c_j | c_\ast, c^\ast)P(E_j | G). \]
	If $G = \Omega$ we simply denote $\bar{u}(a | c_\ast, c^\ast, G)$ as $\bar{u}(a | c_\ast, c^\ast)$
	\end{defi}
	\begin{teo}
	For any bounded decision problem with extreme consequences $c^\ast \succ c_\ast$ and any $G \succ \emptyset$
	\[ a_1 \succeq_G a_2 \Leftrightarrow \bar{u}(a_1) \geq \bar{u}(a_2). \]
	\end{teo}
	\begin{proof}
	See \cite{bernardo2000bayesian} Chapter 2 “Foundations”, Proposition 2.22.
	\end{proof}
	This result means that for any decision maker whose preferences satisfies axioms 1 through 5 the only election criteria that is compatible with the axioms is the maximization of expected utility and thus establishes a \textit{normative} criteria for decision making for rational agents. Of course, this result establishes a complete ordering of the options, but it does not guarantees the existence of an option for which the expected utility is maximum and further mathematical assumptions are required over the utility function $u$ in order to guarantee the existence of the maxmimum utility option (\cite{bernardo2000bayesian}).

	The result extends previous work of \cite{von1944theory} who assumed that the decision maker knows the probabilties of the uncertain events and guarantees the existence of an utility function such that the decision maker chooses by calculating expected value of this utility function.
	\subsection{Game Theory}
	A \textit{game} arises when two or more \textit{rational} decision makers, or players, have to make decisions in a situation in which the outcome for each player is partly determined by the choices of other players (\cite{binmore2008rational}).

	Game Theory is an area of Mathematics which is used to model games. The basic assumptions of the theory is that each decision maker is rational, and they take into account their knowledge or expectations of other decision makers behavior.

	In this section we will review the basic models of Game Theory that are required to this thesis proposal, which are the normal-form game, and the extensive-form game.

	The normal-form game models a situation in which two or more agents interact and where each one of them will choose an action simultaneously. Also, it is assumed that all of the relevant aspects of the game are known for each player.

	\begin{defi}
	A normal-form game is a tuple $(N,A,(\succeq_i)_{i \in N})$ where:
	\begin{itemize}
	\item $N$ is a finite set of players.
	\item $A=A_1 \times ... \times A_n$ where each $A_i$ is the set of available actions for player $i$. Each $a=(a_1,...,a_n) \in A$ is called an action profile.
	\item For each player $i \in N$, a preference relation $\succeq_i$ defined over $A$.
	\end{itemize}
	\end{defi}

Because of the decision-making criteria we know that each player can replace his preferences $\succeq_i$ for a utility function $u_i$.

The prisoner's dilemma is the most common example of a normal-form game, where two suspects must either confess or remain silent about a crime and they can't communicate with each other. In this game, both of the players know the consequences of each \textit{outcome} and the rewards (utilities) associated with each outcome.

In some cases, the players do not know all of the relevant aspects of a game, such as the payoffs or the available actions to other players. It is this situation that is modeled by normal-form Bayesian games (\cite{osborne1994course}, \cite{shoham2008multiagent}).

\begin{defi}
A normal-form Bayesian game consists of:
\begin{itemize}
\item A set of states $\mathcal{S}$.
\item A finite set of players $N$.
\end{itemize}
\end{defi}

\section{Problem Statement}
	\subsection{Problem Statement}
	Let  $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem under uncertainty whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. And consider a rational decision maker who doesn't know the parameters of the causal model which control the probabilities of observing a consequence given an action $a \in \mathcal{A}$, we ask how the decision maker could learn about the causal structure that controls his environment in order to make good choices with respect to the decision makers' preferences.
	\subsection{Motivation}
	Decision making under uncertain conditions is a fundamental part of intelligent reasoning (\cite{lake2017building}). Intelligent agents often face situations where an action must be chosen in the presence of uncertain conditions; this means that an outcome will be observed according to some probability distribution given the action chosen by the agent.

@@

	It is known that human beings conceive their actions on the world as \textit{intervening} in the world (\cite{hagmayer2009decision}). Following this idea, \cite{lattimoreNIPS2016} consider decision problems where the action to be chosen is an intervention over a known causal graphical model. The agent must choose the intervention that maximizes the value of a \textit{target variable} after a series of learning rounds. They model their problem as if choosing an intervention was choosing an arm of a \textit{slot machine}, in which a gambler chooses an arm and gets some reward. From the rewards they estimate probabilities and output an optimal action in the sense of obtaining minimal regret. Their work considers that the causal model is fully known. They mention that the case where the causal model is unknown is left as an open question, and it is precisely what we are proposing to answer. 
	\subsection{Justification}
	Learning a causal model of some phenomena is important, since it allows to \textit{explain} aspects of the model that a purely associative model would not be abel to explain.

	\subsection{Research Question}
	How a rational decision maker who faces an uncertain environment which is governed by a causal mechanism can learn and make use of this causal structure in order to make good choices? How can causal structure help a decision maker in order to guide his learning process?
	\subsection{Hypothesis}
	Let $\mathcal{G}$ a causal graphical model and let $(\mathcal{A},\mathcal{E},\mathcal{C})$ a decision problem whose actions $a_i = \{ c_j | E_j : j \in J \}$  are causally related to consequences $c \in \mathcal{C}$ through the uncertain events $E \in \mathcal{E}$ which correspond to variables in $\mathcal{G}$. Then, if a decision maker doesn't know the probabilities of the uncertain events in $E$, by repeatedly making decisions he can learn causal information and use it to find the optimal actions (in the sense of expected utility theory) in less, or equal, rounds than if he doesn't consider causal information.
	\subsection{Objectives}
		\subsubsection{General Objectives}
		\subsubsection{Specific Objectives}
		\subsubsection{Limitations}

@@

\section{Methodology}
\subsection{Modelling}
\subsection{Working plan}
\subsubsection{Publication plan}
\begin{itemize}
\item Gonzalez-Soto, M., Sucar, L.E., Escalante, H.J., \textbf{Playing against Nature: causal discovery for decision making under uncertainty}. Accepted for a poster presentation at the CausalML 2018 workshop at ICML 2018.
\end{itemize}

\section{Preliminary Results}
To show the factibility of this proposal, we considered a test scenario and two cases in ascending difficulty
\newpage
\bibliographystyle{apalike}
\bibliography{/Users/MauricioGS1/INAOE/Propuesta/Bibliografia.bib}
\end{document}
